---
title: "HEAL EEG Biomarkers"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
---

<style>
img { max-width: none; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

rm(list = ls())
gc()

required_packages <- c(
  "knitr", "ggplot2", "tableone", "corrplot", "PerformanceAnalytics",
  "ggExtra", "EnvStats", "sandwich", "DescTools", "lmtest", "glmnet",
  "mice", "pracma", "dotwhisker", "redcapAPI", "data.table",
  "Hmisc", "dplyr", "writexl", "readxl", "REDCapR", "janitor",
  "ggrepel", "pathfindR", "org.Hs.eg.db"
)

missing_packages <- required_packages[!required_packages %in% rownames(installed.packages())]
if (length(missing_packages) > 0) {
  message("Install missing packages before knitting: ", paste(missing_packages, collapse = ", "))
}

invisible(lapply(required_packages, require, character.only = TRUE))
```

# Notes

This document is saved as HTML because some figures are too wide to fit on a PDF page. If figures do not render well in the browser, download the HTML file and open it locally.

This GitHub version has been sanitized for public sharing:

- REDCap API tokens have been removed.
- Local file paths have been replaced with configurable paths.
- Private identifiers and institution-specific paths should be reviewed before posting.
- Raw data are not included in this repository.

# Configuration

Create a local `.Renviron` file, not committed to GitHub, containing:

```bash
REDCAP_URI="https://redcap.ucsf.edu/api/"
REDCAP_TOKEN="your_redcap_token_here"
```

Set paths below to match your local environment. For GitHub, place de-identified input files under `data/` and generated outputs under `outputs/`.

```{r config}
paths <- list(
  cytokine_file = "data/Cytokines_for_NSR_RISE.xlsx",
  output_dir = "outputs"
)

if (!dir.exists(paths$output_dir)) {
  dir.create(paths$output_dir, recursive = TRUE)
}

redcap_uri <- Sys.getenv("REDCAP_URI")
redcap_token <- Sys.getenv("REDCAP_TOKEN")
```

# Data processing

We identified three samples with unusual biomarker readings that may have been urine samples instead of blood. Those three samples, the second samples for CLA-101, CLA-201, and PIT-113, were removed from this analysis.

For biomarker concentrations reported as thresholds, for example `< some value`, the threshold was used as the value.

These analyses were updated with the re-run tau measurements.

Throughout, biomarkers are ordered and grouped as follows:

- Brain-specific proteins: GFAP, NSE, S100B, UCH-L1, Tau
- Other growth factors: BDNF, EPO, NCAM1/CD56, NRG1β, VEGF
- Pro-inflammatory cytokines, chemokines, and peptides: C5a, Fractalkine/CXCL1, ICAM1/CD54, IFN-γ, IL-1β, IL-6, IL-12p70, IL-17A, MCP-1/CCL2, MIP-1α/CCL3, MIP-1β/CCL4, VCAM1/CD106
- Anti-inflammatory cytokines: IL-1RA, IL-8, IL-10, IL-13, IL-33

```{r import-data, eval=FALSE}
# Option A: Export directly from REDCap.
# Do not hard-code REDCap tokens in this file.

if (redcap_uri == "" || redcap_token == "") {
  stop("Set REDCAP_URI and REDCAP_TOKEN in your local .Renviron file.")
}

rcon <- redcapConnection(url = redcap_uri, token = redcap_token)

derived.data <- exportRecordsTyped(
  rcon,
  fields = NULL,
  forms = NULL,
  records = NULL,
  events = NULL,
  labels = TRUE,
  dates = TRUE,
  survey = TRUE,
  checkboxLabels = FALSE,
  factors = FALSE
)

cytokines <- readxl::read_excel(paths$cytokine_file, na = ".")
derived.data <- merge(derived.data, cytokines, by = "study_id", all = TRUE)
```

```{r import-from-local, eval=FALSE}
# Option B: Load a de-identified RDS exported from REDCap.
# This is preferred for public GitHub workflows.

derived.data <- readRDS("data/derived_data_deidentified.rds")
cytokines <- readxl::read_excel(paths$cytokine_file, na = ".")
derived.data <- merge(derived.data, cytokines, by = "study_id", all = TRUE)
```

# Variable derivation

```{r variable-derivation, eval=FALSE}
# Preserve one study ID with missing blood volume that was known to be valid.
derived.data.extra <- subset(derived.data, study_id == "02308")
derived.data <- subset(derived.data, vol_blood != "NA")
derived.data <- bind_rows(derived.data, derived.data.extra)

# Exclude participants excluded at 24 months.
derived.data <- subset(derived.data, final_24_excluded != "Yes" | is.na(final_24_excluded))

# Participant type.
derived.data$patient_type <- as.factor(derived.data$patient_type)
levels(derived.data$patient_type) <- list("Cases" = c("Acute"), "Controls" = c("Control"))

# Demographics and clinical variables for Table 1.
derived.data$`Maternal race` <- as.factor(derived.data$race)
levels(derived.data$`Maternal race`) <- list(
  "White" = c("White"),
  "Black" = c("Black / African American"),
  "Asian" = c("Asian"),
  "Multiple/Other/Unknown" = c(
    "American Indian / Alaska Native",
    "Native Hawaiian / Other Pacific Islander",
    "Other",
    "Moret than one race",
    "Unknown / Not Reported",
    "Decline to Answer"
  )
)

derived.data$`Maternal ethnicity` <- as.factor(derived.data$ethnicity)
levels(derived.data$`Maternal ethnicity`) <- list(
  "No/not reported" = c("NOT Hispanic or Latino", "Unknown / Not reported"),
  "Yes" = c("Hispanic or Latino")
)

derived.data$`Maternal education` <- as.factor(derived.data$maternal_edu)
levels(derived.data$`Maternal education`) <- list(
  "At least some college or not reported" = c("College graduate", "Graduate study", "Unknown/Unavailable", "Decline to Answer"),
  "High school or less" = c("High school graduate", "Some education, high school not complete")
)

derived.data$`Insurance Type` <- as.factor(derived.data$insurance_type)
levels(derived.data$`Insurance Type`) <- list(
  "Private" = c("Private"),
  "Public" = c("Public", "Unknown", "Decline to Answer", "None")
)

derived.data <- derived.data %>%
  mutate(`Maternal chorioamnionitis` = case_when(
    maternal_fever == "No Fever" ~ "No",
    maternal_fever %in% c(
      "Fever > 39C (102.2F) once",
      "Fever > 38C (100.4F) to 38.9C (102.0F) on two occasions"
    ) ~ "Yes",
    TRUE ~ NA_character_
  ))

derived.data$`Cesarean section delivery` <- as.factor(derived.data$delivery_mode)
levels(derived.data$`Cesarean section delivery`) <- list(
  "No" = c("Vaginal"),
  "Yes" = c("Operative vaginal (vacuum or forceps)", "Scheduled cesarean section", "Emergent cesarean section")
)

derived.data$`Birth location` <- as.factor(derived.data$location)
levels(derived.data$`Birth location`) <- list(
  "Inborn" = c("Study center inpatient ward or ICU", "Study center L&D or well baby nursery"),
  "Other hospital/other" = c("Referral center L&D, well baby nursery, or ICU", "Home")
)

derived.data$Sex <- as.factor(derived.data$gender)
levels(derived.data$Sex) <- list("Male" = c("Male"), "Female" = c("Female"))

derived.data$`Birth weight, g` <- as.numeric(derived.data$birth_weight)
derived.data$`Gestational age at birth, wk` <- as.numeric(derived.data$gest_age)
derived.data$`5 min Apgar` <- as.numeric(derived.data$apgar_5min)
derived.data$`10 min Apgar` <- as.numeric(derived.data$apgar_10min)
derived.data$`Gestational age at neonatal blood draw, wk` <- as.numeric(derived.data$postmens_age_neobd)
```

# Biomarker naming and ordering

```{r biomarker-names, eval=FALSE}
# Rename raw biomarker columns to analysis-friendly names.
# Keep this block synchronized with the input assay file.

data.table::setnames(
  derived.data,
  old = c(
    'hGFAP1','Tau1','Fractalkine1','IFNy1','IL-101','IL-12(p70)1','IL-131','IL-17A1','IL-1b1',
    'IL-211','IL-231','IL-41','IL-51','IL-61','IL-81','MIP-1a1','MIP-1b1','TNFa1','BDNF1',
    'CCL2/JE/MCP-11','Complement Component C5a1','Enolase 2/Neuron-specific Enolase1','Growth Hormone1',
    'ICAM-1/CD541','IL-1ra/IL-1F31','IL-331','NCAM-1/CD561','Neuregulin-1 beta 1/NRG1 beta 11',
    'S100B1','UCH-L1/PGP9.51','VCAM-1/CD1061','VEGF1'
  ),
  new = c(
    'GFAP1','TAU1','CX3CL1/Fractalkine1','IFN-g1','IL-101','IL-12p701','IL-131','IL-171','IL-1b1',
    'IL-211','IL-231','IL-41','IL-51','IL-61','IL-81','CCL3/MIP-1a1','CCL4/MIP-1b1','TNF-a1',
    'BDNF1','CCL2/MCP11','C5a1','NSE1','GH1','CD54/ICAM-11','IL-1RA1','IL-331','CD56/NCAM11',
    'NGB1','s100b1','UCHL11','CD106/VCAM-11','VEGF1'
  ),
  skip_absent = TRUE
)

biomarker_names_brain_specific_proteins <- c("GFAP1", "s100b1", "UCHL11", "TAU1")
biomarker_names_other_growth_factors <- c("BDNF1", "CD56/NCAM11", "GH1", "NGB1", "NSE1", "VEGF1")
biomarker_names_pro_inflammatory <- c(
  "C5a1", "CX3CL1/Fractalkine1", "CD54/ICAM-11", "IFN-g1", "IL-1b1", "IL-41", "IL-51",
  "IL-61", "IL-81", "IL-12p701", "IL-171", "IL-211", "IL-231", "CCL2/MCP11", "CCL3/MIP-1a1",
  "CCL4/MIP-1b1", "CD106/VCAM-11", "TNF-a1"
)
biomarker_names_anti_inflammatory <- c("IL-1RA1", "IL-101", "IL-131", "IL-331")

biomarker_names <- c(
  biomarker_names_anti_inflammatory,
  biomarker_names_pro_inflammatory,
  biomarker_names_brain_specific_proteins,
  biomarker_names_other_growth_factors
)

biomarker_names_core <- unique(substr(biomarker_names, 1, nchar(biomarker_names) - 1))
```

# Cohort tables

```{r cohort-tables, eval=FALSE}
table_3_vars <- c(
  "Maternal race", "Maternal ethnicity", "Maternal education", "Insurance Type",
  "Maternal chorioamnionitis", "Cesarean section delivery", "Birth location", "Sex",
  "Birth weight, g", "Gestational age at birth, wk", "Gestational age at neonatal blood draw, wk",
  "5 min Apgar", "10 min Apgar", "Seizure Etiology", "Worst EEG Background",
  "Days of EEG Seizures", "Therapeutic Hypothermia", "Abnormal Neurologic Exam at Discharge",
  "CP", "WIDEA", "Services"
)

analysis.data <- derived.data %>%
  filter(!is.na(Epilepsy))

table_3 <- CreateTableOne(
  data = analysis.data,
  vars = table_3_vars,
  strata = "Epilepsy",
  test = TRUE,
  addOverall = TRUE
)

table_3_print <- print(
  table_3,
  showAllLevels = FALSE,
  missing = FALSE,
  nonnormal = c("5 min Apgar", "10 min Apgar", "Gestational age at neonatal blood draw, wk")
)

knitr::kable(table_3_print, caption = "Cohort characteristics by epilepsy outcome")
```

# Missing data

```{r missing-data, eval=FALSE}
biomarker.data$number_biomarker <- rowSums(!is.na(biomarker.data[, biomarker_names]))

table_missing_people <- data.frame(table(biomarker.data$number_biomarker))
colnames(table_missing_people) <- c("Number of measures", "Number of participants")
knitr::kable(table_missing_people, caption = "Number of biomarker measures available for participants")
```

# Primary biomarker analysis

```{r primary-biomarker-analysis, eval=FALSE}
manual_winsorize <- function(x, lower = 0.025, upper = 0.975) {
  qnt <- quantile(x, probs = c(lower, upper), na.rm = TRUE)
  ifelse(x < qnt[1], qnt[1], ifelse(x > qnt[2], qnt[2], x))
}

analysis.data$epilepsy <- ifelse(analysis.data$Epilepsy == 1, 1, 0)

primary_df <- data.frame(
  biomarker = biomarker_names,
  median_no = NA_real_,
  iqr_no = NA_real_,
  median_yes = NA_real_,
  iqr_yes = NA_real_,
  rr = NA_real_,
  rr_ci_lower = NA_real_,
  rr_ci_upper = NA_real_,
  rr_p = NA_real_
)

primary_df$biomarker_name <- substr(primary_df$biomarker, 1, nchar(primary_df$biomarker) - 1)
primary_df$sample <- substr(primary_df$biomarker, nchar(primary_df$biomarker), nchar(primary_df$biomarker))
cutoff <- qnorm(1.95 / 2)

for (i in seq_len(nrow(primary_df))) {
  biomarker.i <- primary_df$biomarker[i]

  primary_df$median_no[i] <- median(analysis.data[analysis.data$epilepsy == 0, biomarker.i], na.rm = TRUE)
  primary_df$iqr_no[i] <- IQR(analysis.data[analysis.data$epilepsy == 0, biomarker.i], na.rm = TRUE)
  primary_df$median_yes[i] <- median(analysis.data[analysis.data$epilepsy == 1, biomarker.i], na.rm = TRUE)
  primary_df$iqr_yes[i] <- IQR(analysis.data[analysis.data$epilepsy == 1, biomarker.i], na.rm = TRUE)

  analysis.data$predictor.i <- log2(analysis.data[, biomarker.i])
  model.i <- glm(epilepsy ~ predictor.i, family = poisson, data = analysis.data)

  cov.model.i <- sandwich::vcovHC(model.i, type = "HC0")
  std.err.i <- sqrt(diag(cov.model.i))["predictor.i"]

  primary_df$rr[i] <- exp(coef(model.i)["predictor.i"])
  primary_df$rr_ci_lower[i] <- exp(coef(model.i)["predictor.i"] - cutoff * std.err.i)
  primary_df$rr_ci_upper[i] <- exp(coef(model.i)["predictor.i"] + cutoff * std.err.i)
  primary_df$rr_p[i] <- lmtest::coeftest(model.i, vcov = cov.model.i)["predictor.i", "Pr(>|z|)"]
}

primary_df <- primary_df %>%
  mutate(
    p_fdr = p.adjust(rr_p, method = "fdr"),
    fold_change = median_yes / median_no,
    log2_fc = log2(fold_change)
  )

knitr::kable(primary_df, caption = "Primary biomarker associations with epilepsy")
```

# Manhattan plot

```{r manhattan-plot, eval=FALSE}
primary_df <- primary_df %>%
  mutate(
    category = case_when(
      biomarker_name %in% substr(biomarker_names_brain_specific_proteins, 1, nchar(biomarker_names_brain_specific_proteins) - 1) ~ "Brain-specific proteins",
      biomarker_name %in% substr(biomarker_names_other_growth_factors, 1, nchar(biomarker_names_other_growth_factors) - 1) ~ "Other growth factors",
      biomarker_name %in% substr(biomarker_names_pro_inflammatory, 1, nchar(biomarker_names_pro_inflammatory) - 1) ~ "Pro-inflammatory cytokines, chemokines, and peptides",
      biomarker_name %in% substr(biomarker_names_anti_inflammatory, 1, nchar(biomarker_names_anti_inflammatory) - 1) ~ "Anti-inflammatory cytokines",
      TRUE ~ "Other"
    ),
    neg_log10_fdr = -log10(p_fdr)
  )

manhattan_plot <- ggplot(primary_df, aes(x = factor(biomarker_name, levels = unique(biomarker_name)), y = neg_log10_fdr, color = category)) +
  geom_point(stroke = 1.5) +
  theme_bw() +
  theme(
    text = element_text(size = 16),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title.y = element_text(size = 18.5, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(y = expression("FDR-adjusted p-value " ~ (-log[10] * " scale")), x = "")

manhattan_plot

ggsave(
  filename = file.path(paths$output_dir, "manhattan_plot_RISE.png"),
  plot = manhattan_plot,
  width = 12,
  height = 6,
  dpi = 600
)
```

# Volcano plot

```{r volcano-plot, eval=FALSE}
volcano_df <- primary_df %>%
  mutate(
    gene_type = case_when(
      fold_change >= 1.24 & p_fdr <= 0.05 ~ "up",
      fold_change <= 0.75 & p_fdr <= 0.05 ~ "down",
      TRUE ~ "ns"
    ),
    delabel = if_else(gene_type != "ns", biomarker_name, NA_character_)
  )

volcano_plot <- ggplot(volcano_df, aes(x = log2_fc, y = -log10(p_fdr), fill = gene_type, label = delabel)) +
  geom_point(shape = 21, color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = c(log2(0.75), log2(1.25)), linetype = "dashed", linewidth = 0.2) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = Inf) +
  theme_bw() +
  labs(x = expression(log[2]("fold change")), y = expression("FDR-adjusted p-value " ~ (-log[10] * " scale")))

volcano_plot

ggsave(
  filename = file.path(paths$output_dir, "volcano_RISE.png"),
  plot = volcano_plot,
  width = 6,
  height = 3,
  dpi = 600
)
```

# Protein pathway analysis

```{r protein-pathway-analysis, eval=FALSE}
pathway_input <- primary_df %>%
  transmute(
    Gene_symbol = biomarker_name,
    log2FC = log2_fc,
    P.Value = rr_p
  )

pathway_output <- pathfindR::run_pathfindR(pathway_input, pin_name_path = "KEGG")
pathfindR::enrichment_chart(pathway_output, top_terms = 10)
```

# Session information

```{r session-info}
sessionInfo()
```
